Generalized Bradley-Terry models

### 2025-04-22 inst/jss_hyper3_real.Rnw
better make it 4 decimal places

We now apply the ideas above to a dataset obtained from the {\tt
  preflib} library~\citep{MaWa13a} in which ten different types of
sushi were placed in order of preference by 5000 judges.  Three of the
sushis (maguro, toro, tekka-maki) were made with tuna, and it is
reasonable to view these three types as being exchangeable in much the
same way as the red and blue buses.  To proceed, we need to define
equivalence classes of sushi---one for tuna-based sushis and one for
everything else.  Of course, one can also imagine the opposite
phenomenon whereby the judge decides he has had enough tuna for the
moment and this disincentivises further tuna choices.  This mechanism
would be modelled by having $\lambda<1$.  Implementation in
\code{hyper3} formalism is straightforward.  The dataset used is
\code{sushi\_table}, part of the \pkg{hyper2} package:
%
<<showsushitable>>=
noquote(head(sushi_table))
@
%
Thus, from the top line, we see that the first judge placed tamago
first, followed by anago and ika.  The equivalence classes are given
by
%
<<show_sushi_eq_classes>>=
sushi_eq_classes
@
%
so class 2 is the tuna group and class 1 everything else.  We can now
define a bespoke likelihood function for these observations, using
$\lambda$ as the argument:
%
<<defcalculate_sushi_H>>=
make_sushi_H <- function(lambda){
    H <- hyper3()
    for(i in seq_len(nrow(sushi_table))){
        H <- H + cheering3(sushi_table[i,], e = sushi_eq_classes, h = c(1, lambda))
    }
    return(H)
}
@
%
<<calclikesushi, echo = FALSE, cache = TRUE>>=
f <- function(lambda){
   H <- make_sushi_H(lambda)
   maxp(H, give = 1)$`Log-likelihood`
}
s <- exp(seq(from = log(0.9), to = log(10), len = howmany))
L <- sapply(s, f)
@ 
%
% chunk 'findstuffsushi' takes about 10-20 minutes to run
%
<<findstuffsushi, echo = FALSE, cache = TRUE>>=
osushi <- optimize(f, c(1, 4), maximum = TRUE)
@ 
%
% chunk tinysushi' takes about 5 minutes to run
%
<<tinysushi, echo = FALSE, cache = TRUE>>=
sushimaxlambda <- osushi$maximum
sushisupport <- osushi$objective - f(1)
@
%
% chunk 'findotherstuffsushi' takes about 10 minutes to run
%
<<findothersushistuff, echo = FALSE, cache = TRUE>>=
g <- function(lambda){f(lambda) - osushi$objective + 2}
sushi_lower <- uniroot(g, interval = exp(c(0, 1)), tol = 0.01)$root
sushi_upper <- uniroot(g, interval = exp(c(1, 2)), tol = 0.01)$root
@ 
%
(above, we note that the return value {\tt H} is the sum of successive
log-likelihood functions of each row of the table, assumed to be
independent).  Then, as before, we may define a profile likelihood
function of $\lambda$, Figure~\ref{profilesupportsushi}.  We
reject~$H_0\colon\lambda=1$ in favour of
$H_A\colon\lambda=\Sexpr{round(sushimaxlambda,2)}$ with about
$\Sexpr{round(sushisupport,2)}$ units of support.  Alternatively, we
may observe that this lies in the tail region of its asymptotic
$\chi^2_1$ distribution with a $p$~value of about
$\Sexpr{round(pchisq(2*sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
therefore be reasonably sure that tuna sushi are somehow analagous to
red and blue buses, in the sense that all three tuna items tend to
occupy closely separated places in judges' preference ranks.

>>>>

We now apply the ideas above to a dataset obtained from the {\tt
  preflib} library~\citep{MaWa13a} in which ten different types of
sushi were placed in order of preference by 5000 judges.  Three of the
sushis (maguro, toro, tekka-maki) were made with tuna, and it is
reasonable to view these three types as being exchangeable in much the
same way as the red and blue buses.  To proceed, we need to define
equivalence classes of sushi---one for tuna-based sushis and one for
everything else.  Of course, one can also imagine the opposite
phenomenon whereby the judge decides he has had enough tuna for the
moment and this disincentivises further tuna choices.  This mechanism
would be modelled by having $\lambda<1$.  Implementation in
\code{hyper3} formalism is straightforward.  The dataset used is
\code{sushi\_table}, part of the \pkg{hyper2} package:
%
<<showsushitable>>=
noquote(head(sushi_table))
@
%
Thus, from the top line, we see that the first judge placed tamago
first, followed by anago and ika.  The equivalence classes are given
by
%
<<show_sushi_eq_classes>>=
sushi_eq_classes
@
%
so class 2 is the tuna group and class 1 everything else.  We can now
define a bespoke likelihood function for these observations, using
$\lambda$ as the argument:
%
<<defcalculate_sushi_H>>=
make_sushi_H <- function(lambda){
    H <- hyper3()
    for(i in seq_len(nrow(sushi_table))){
        H <- H + cheering3(sushi_table[i,], e = sushi_eq_classes, h = c(1, lambda))
    }
    return(H)
}
@
%
<<calclikesushi, echo = FALSE, cache = TRUE>>=
f <- function(lambda){
   H <- make_sushi_H(lambda)
   maxp(H, give = 1)$`Log-likelihood`
}
s <- exp(seq(from = log(0.9), to = log(10), len = howmany))
L <- sapply(s, f)
@ 
%
% chunk 'findstuffsushi' takes about 10-20 minutes to run
%
<<findstuffsushi, echo = FALSE, cache = TRUE>>=
osushi <- optimize(f, c(1, 4), maximum = TRUE)
@ 
%
% chunk tinysushi' takes about 5 minutes to run
%
<<tinysushi, echo = FALSE, cache = TRUE>>=
sushimaxlambda <- osushi$maximum
sushisupport <- osushi$objective - f(1)
@
%
% chunk 'findotherstuffsushi' takes about 10 minutes to run
%
<<findothersushistuff, echo = FALSE, cache = TRUE>>=
g <- function(lambda){f(lambda) - osushi$objective + 2}
sushi_lower <- uniroot(g, interval = exp(c(0, 1)), tol = 0.01)$root
sushi_upper <- uniroot(g, interval = exp(c(1, 2)), tol = 0.01)$root
@ 
%
(above, we note that the return value {\tt H} is the sum of successive
log-likelihood functions of each row of the table, assumed to be
independent).  Then, as before, we may define a profile likelihood
function of $\lambda$, Figure~\ref{profilesupportsushi}.  We
reject~$H_0\colon\lambda=1$ in favour of
$H_A\colon\lambda=\Sexpr{round(sushimaxlambda,2)}$ with about
$\Sexpr{round(sushisupport,2)}$ units of support.  Alternatively, we
may observe that this lies in the tail region of its asymptotic
$\chi^2_1$ distribution with a $p$~value of about
$\Sexpr{round(pchisq(2*sushisupport,df=1,lower.tail=FALSE),4)}$.  We can
therefore be reasonably sure that tuna sushi are somehow analagous to
red and blue buses, in the sense that all three tuna items tend to
occupy closely separated places in judges' preference ranks.
